Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-03-27 talk/icooolps2011/paper.tex
second dynamo ref
few fixes, comment out a section to save space

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left you
\pedronis{XXX style: do we want to use you, or we and one} see the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives you ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation.

@@

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is a pure function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{purefunction} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If you analyze this function, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function from the outside is pure. This is
a property that is not easily detectable by analysis. Therefore, the purity
of this function needs to be annotated.

@@

\subsection{More General Patterns}

The techniques we used above to make instance and class lookups faster are
applicable in more general cases than the one we developed them for. A more
abstract view of maps is that of splitting a data-structure into a part that
changes slowly, and a part that changes quickly. In the concrete example of maps
we split the original dictionary into the map (the slow-changing part) and the
storage array (the quick-changing part). All the computation on the
slow-changing part can be constant-folded during tracing so that only the
manipulation of the quick-changing part remains.

Similarly, versions can be used to constant-fold arbitrary functions of large data
structures. The version needs to be updated carefully every time the result of
this function can change. Therefore this is useful only if the data structure is
expected to change slowly.

@@

The very first meta-tracer is described by Sullivan et. al.
\cite{sullivan_dynamic_2003}. They used Dynamo RIO, the successor of Dynamo
\cite{XXX} to trace through a small synthetic interpreter. As in Dynamo, tracing
happens on the machine code level. The tracer is instructed by some hints in the
tiny interpreter where the main interpreter loop is and for how long to trace to
match loops in the user-level functions. These hints are comparable to the one
PyPy uses for the same reasons \cite{bolz_tracing_2009}. Their approach suffers
mostly from the low abstraction level that machine code provides.

@@

An early attempt at building a general environment for implementing languages
efficiently is described by Wolczko et. al. \cite{mario_wolczko_towards_1999}.
They implement Java and Smalltalk on top of the SELF VM by compiling the
languages to SELF. The SELF JIT is good enough to optimize the compiled code
very well. We believe the approach to be restricted to languages that are
similar enough to SELF. XXX

@@

is there anything about versions? smalltalks tend to clear their method caches
when new methods are added. self and java use dependency tracking and
deoptimization. this is better what we have above, because we need runtime
checks. mention out of line guard?

>>>>

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left are
the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

@@

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation.

@@

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is a pure function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{purefunction} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If this function is analyzed, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function from the outside is pure. This is
a property that is not easily detectable by analysis. Therefore, the purity
of this function needs to be annotated.

@@

%\subsection{More General Patterns}
%
%The techniques we used above to make instance and class lookups faster are
%applicable in more general cases than the one we developed them for. A more
%abstract view of maps is that of splitting a data-structure into a part that
%changes slowly, and a part that changes quickly. In the concrete example of maps
%we split the original dictionary into the map (the slow-changing part) and the
%storage array (the quick-changing part). All the computation on the
%slow-changing part can be constant-folded during tracing so that only the
%manipulation of the quick-changing part remains.
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.

@@

The very first meta-tracer is described by Sullivan et. al.
\cite{sullivan_dynamic_2003}. They used Dynamo RIO, the successor of Dynamo
\cite{bala_dynamo:_2000} to trace through a small synthetic interpreter. As in Dynamo, tracing
happens on the machine code level. The tracer is instructed by some hints in the
tiny interpreter where the main interpreter loop is and for how long to trace to
match loops in the user-level functions. These hints are comparable to the one
PyPy uses for the same reasons \cite{bolz_tracing_2009}. Their approach suffers
mostly from the low abstraction level that machine code provides.

@@

An early attempt at building a general environment for implementing languages
efficiently is described by Wolczko et. al. \cite{mario_wolczko_towards_1999}.
They implement Java and Smalltalk on top of the SELF VM by compiling the
languages to SELF. The SELF JIT is good enough to optimize the compiled code
very well. We believe the approach to be restricted to languages that are
similar enough to SELF as there were no mechanisms to control the underlying
compiler.

@@

%is there anything about versions? smalltalks tend to clear their method caches
%when new methods are added. self and java use dependency tracking and
%deoptimization. this is better what we have above, because we need runtime
%checks. mention out of line guard?
